Species-specific effects of the Urban Heat Island on tree growth across Berlin

### 2021-09-15 analysis/paper/manuscript.Rmd
updating ms

Berlin is one of the largest metropolitan areas in Central Europe (892$~km^2$) with a population of approximately 3.6 million, and a maximum extent of 38$~km$ in North-South and 45$~km$ in East-West directions. 
It is located in North-Eastern Germany, and lies in the temperate zone with warm-humid climate (Dfb) according to the updated Köppen-Geiger classification [@beck2018], with mean annual temperature of approximately 10$^\circ C$ and precipitation of 575$~mm$ (Tempelhof weather station, DWD).
Berlin features low relief (approximately 30$~m$ to 60$~m$ with 120$~m$ at solitary peaks), and is centered around a glacial outwash valley (sands, gravel), bordered by two plateaus consisting of glacial till and clay in the North-East and South, as well as sands in the South-West.
The city provides extensive public green space covering around 30$~\%$ of its area [@suvkberlin2019], with an extensive urban forest of nearly 700000 publically-managed trees along streets, in parks and in riparian areas.
It has a pronounced urban heat island, owing to its high degree of built-up and sealed surfaces, with air temperatures up to 12$^\circ C$ higher in urban compared to adjacent rural areas [@fenner2014].

## Data sources

@@

@@

# Preliminary results {#sec:results}

@@

1. Temperature, environmental and urban controls on tree growth  
	+ Intra specific differences
	+ Sensitive trees in hot spots
	+ Temporal dynamics  
	+ Validity and limitations

2. Implications:  
  + Policy / planning:  
    - tree species
	  - Increase temporal coverage and accuracy  
		- which trees to prioritize for measurement campaigns  
		- Supplement species matrix etc. with these analyses  
  + Open data important --> drives innovation

# Outlook

We seek to build upon and improve the current analysis by:

- validating the database with independent observations
- incorporating more pertinent covariates as dependent variables in the linear mixed model  
- testing multiple model structures with formal model selection procedures
- checking model residuals for spatial auto-correlation and accounting for it where necessary to ensure unbiased estimates of effect sizes
- repeating the above with a hierarchical GAM (i.e. GAMM) to allow for:  
  * estimating continuous prediction surfaces for UHI impacts on individual species' growth (similar to results in Figure$~$\@ref(fig:fig-lmestat)) under recent conditions
  * estimating absolute, species-specific growth potential under increased temperatures and UHI loading under climate change, ideally based on climate simulations (otherwise step-wise increases based on RCP scenarios) for the key species. Note, that  complications presented by 'out-of-sample' predictions will be addressed.
  * assess potential age-dependent UHI impacts on individual species.  
  * repeating the above (GAMM) with total and incremental basal area as responses

>>>>

Berlin is one of the largest metropolitan areas in Central Europe (892$~km^2$) with a population of approximately 3.6 million, and a maximum extent of 38$~km$ in North-South and 45$~km$ in East-West directions. 
It is located in North-Eastern Germany, and lies in the temperate zone with warm-humid climate (Dfb) according to the updated Köppen-Geiger classification [@beck2018], with mean annual temperature of approximately 10$^\circ C$ and precipitation of 575$~mm$ (Tempelhof weather station, German Weather Service DWD).
Berlin has low relief (approximately 30$~m$ to 60$~m$ with 120$~m$ at solitary peaks), and is centered around a glacial outwash valley (sands, gravel), bordered by two plateaus consisting of glacial till and clay in the North-East and South, as well as sands in the South-West.
The city provides extensive public green space covering around 30$~\%$ of its area [@suvkberlin2019], with an extensive urban forest of nearly 700000 publically-managed trees along streets, in parks and in riparian areas.
It has a pronounced urban heat island, owing to its high degree of built-up and sealed surfaces, with air temperatures up to 12$^\circ C$ higher in urban compared to adjacent rural areas at night times [@fenner2014].

# General approach

- Assess differences in mean growth across age for individual species produced by environmental factors, including air temperature (or measure of urban heat island loading), soil characteristics and rooting area, adjacent building height) in a space-for-time approach
- Focus on street trees where species have a minimum of X individuals
- Account for potential differences in tree management by authorities
- Use flexible models that can produce curves of varying modality 
- Understand / confirm urban tree growth (assessed for single species)

## Data sources and processing

@@

- datacleanr
- adjusting bad records
- final original, final data set available online, with cleaning scripts

@@

# Results {#sec:results}

@@

1. Temperature, environmental and urban controls on tree growth  
	+ Urban tree growth accelerated (BIWI data)
	+ Intra specific differences in UHI response (note temp. products)
	+ Sensitive trees in hot spots
	+ Other controls: soil, urban fabric, 
	+ Validity and limitations (spatial auto-correlation, data quality, coverage)

2. Implications:  
  + Policy / planning:  
    - tree species
	  - Increase temporal coverage and accuracy of surveys
		- which trees to prioritize for measurement campaigns (low-abundance species, highly sensitive)
		- Compare and Supplement species matrix etc. with these analyses  
		- Generate co-located environmental observations (e.g., micro-climate)
  + Open data important --> drives innovation

# Conclusion
